\documentclass[12pt]{article}
\input{Preamble}

\DeclareGraphicsRule{*}{mps}{*}{}
\usepackage{feynmp}

\begin{document}
~
\vspace{-2cm}
\begin{center}
  {\Huge Homework \#6}
\end{center}
\vspace{1cm}

\noindent Please hand in by next Thursday (November 27), either during
the lecture or email to \texttt{<Philipp.Kleinert@physics.ox.ac.uk>}.
Tutorials location: Seminar Room (21. November), Conference Room
(28. November).
\bigskip

\noindent
In this homework we compute the diagram
\begin{center}
  {\fmfframe(8,15)(8,15){
      {
        \begin{fmffile}{Homework-6-fish}
          \begin{fmfgraph*}(170,20)
            \fmfleft{x1,x2}
            \fmfright{x3,x4}
            \fmf{plain}{x1,left}
            \fmf{plain}{x2,left}
            \fmf{plain}{x3,right}
            \fmf{plain}{x4,right}
            \fmf{plain,left}{left,right,left}
            \fmfdot{left,right}
            \fmflabel{$k_1$}{x1}
            \fmflabel{$k_2$}{x2}
            \fmflabel{$k_3$}{x3}
            \fmflabel{$k_4$}{x4}
          \end{fmfgraph*}
        \end{fmffile}}
    }}
\end{center}
in 4-d and dimensional regularization.
\begin{enumerate}
\item Write the coupling constant as $\lambda_0$ times a suitable
  power of the mass scale $M$ in $4-\omega$ dimensions.
\item Show that the Feynman rules in momentum space and after Wick
  rotation yield
  \begin{equation}
    \nonumber
    A(\omega) = \frac{\lambda_0^2}{2} 
    M^{2\omega} \int \frac{d^{4-\omega}\ell}{(2\pi)^{4-\omega}}
    \frac{1}{\ell^2+m^2} \frac{1}{(\ell-k_1-k_2)^2+m^2}
  \end{equation}
  for the amputated diagram.
\item Use Feynman's identity
  \begin{equation}
    \nonumber
    \frac{1}{a_1\cdots a_n} = 
    \frac{1}{(n-1)!} \int_{\{x_j\geq 0\}} 
    \frac{
      (\prod dx_j) ~ \delta(\sum x_j - 1)
    }{
      (a_1x_1 + \cdots + a_n x_n)^n
    }
  \end{equation}
  to rewrite the loop integration as
  \begin{equation}
    \nonumber
    A(\omega) = \frac{\lambda_0^2}{2} 
    M^{2\omega} \int_0^1 dx \int 
    \frac{d^{4-\omega}p}{(2\pi)^{4-\omega}}
    \frac{1}{\big(p^2 + x(1-x)(k_1+k_2)^2+m^2\big)^2}
  \end{equation}
\item Show that the momentum integral evaluates to
  \begin{equation}
    \nonumber
    A(\omega) = 
    M^{\omega} \frac{\lambda_0^2}{32\pi^2} 
    \left[
      \frac{2}{\omega} - \gamma - 
      \int_0^1 \!\!dx \; 
      \ln\left(\frac{m^2+(k_1+k_2)^2 x (1-x)}{4\pi M^2}\right)
      + O(\omega)
    \right]
  \end{equation}
\item Unlike the tadpole, the finite part of $A(\omega)$ depends on
  the external momenta. In the four-point function, this diagram
  appears three times with permutations of $k_1$, $\dots$, $k_4$. Draw
  the three graphs. Is their $\omega$-independent part of the sum
  $A(\omega; k_1, k_2) + A(\omega; k_1, k_3) + A(\omega; k_1, k_4)$ an
  analytic function of the external momenta? Hint: We are still in
  Euclidean space.
\end{enumerate}

\end{document}
